---
title: "Data exploration with `tidyverse`"
author: '[Kevin Y. X. Wang](http://www.maths.usyd.edu.au/ut/people?who=KY_Wang&sms=y)'
date: "10 May 2018"
output:
  ioslides_presentation:
    fig_height: 3
    fig_width: 4
    incremental: yes
    self_contained: yes
    transition: faster
    widescreen: yes
    smaller: true
always_allow_html: yes
---

```{r setup, include=FALSE}
library(tidyverse)


options(tibble.print_max = 10, tibble.print_min = 5)
```



<style>
slides > slide { overflow: scroll; }
slides > slide:not(.nobackground):after {
content: '';
}
</style>




## Preparing for this lecture {.build}

+ All materials are on https://github.com/kevinwang09/MSH4_2018 and https://github.com/kevinwang09/2017_STAT3914. 

+ Some cheatsheets: https://www.rstudio.com/resources/cheatsheets/.

+ There is an interactive tutorial on `ggplot2`. Just run the following code. 

```{r, eval = F}
## Might be a while...

install.packages("tidyverse") 
install.packages(c("ggplot2","dplyr", "readr","tidyr","janitor","plotly",
                   "devtools","learnr","gapminder", "e1071")) 

library(devtools)
install_github("kevinwang09/2017_STAT3914", subdir = "learnr3914")
```



## Why am I here? 

+ ~~I served at the commands of Jean~~.

+ Everything I am telling you today has an equivalent solution in base `R`. 

+ So why am I even here? 


## What is the aim of this lecture? {.build}

+ Certain tasks in `R` used to be very difficult. Especially data cleaning and manipulations. 

+ For example, how would you calculate the mean of `Sepal.Length`, `Sepal.Width`, `Petal.Length`, `Petal.Width` in the `iris` dataset?

```{r}
head(iris, 3)
mean(iris$Sepal.Length)
mean(iris$Sepal.Width)
```

+ I am too lazy to type... But I think you get the idea...

## What about the grouped means? {.build}

+ Dear me... This is too much typing!!

```{r}
tapply(iris$Sepal.Length, iris$Species, mean)
tapply(iris$Sepal.Width, iris$Species, mean)
```


## You are not alone in hating all these typing! {.build}

+ The `tidyverse` is a coherent collection of packages for data manipulation, exploration and visualisation that share a common design philosophy.

1.  Data should be managed in `data.frame`.
2.  The `data.frame` should be easy to manipulate.
3.  Each row should be an observation and each column should be a column. 

+ The `tidyverse` solution for calculating group means: the process of calculating mean is identical. 

+ All we need to do is to apply this process to each `numeric` column of `iris`. 
```{r}
iris %>%
  summarise_if(is.numeric, mean)
```

+ To caluclate grouped mean: identical to before, but we need to apply the mean to the `Species` grouping first! 

```{r}
iris %>%
  group_by(Species) %>%
  summarise_if(is.numeric, mean)
```



## `tidyverse` is a collection of packages {.build}

+ Reading in data using `readr` and `readxl`
+ Better than data.frame: `tibble` 
+ Basic data cleaning using `janitor`
+ Clean coding using `magrittr`
+ Data filtering using `dplyr`
+ Data visualisation using `ggplot2`


# Reading in data

## `readr` for basic data types and `readxl` for .xlsx files {.build}

+ I know some of you are using this already.
+ Parsing warnings.
+ NEVER read in factors.

```{r}
rawEyData = read_csv("EY-data-science-challenge-pack/train.csv",
                     guess_max = 10, ## use the first 10 rows to guess the variable type
                     progress = TRUE ## If the data is large, a progress bar will be shown
)
```


# tibble


## `tibble` is more advanced data.frame

+ There is a super advanced way of using `tibble`, see another set of slides. 

```{r}
class(rawEyData)
rawEyData
```


# Janitor

## Consistent naming of columns

+ This is perhaps the most useful function in `janitor`.

```{r}
eyData = janitor::clean_names(rawEyData)
eyData
```

## Tabulate in style

+ Sometimes you want to calculate counts as well as percentages. 

```{r}
eyData %>% 
  janitor::tabyl(location, gender) %>% 
  janitor::adorn_percentages() %>% 
  janitor::adorn_ns()
```



# Magrittr


## `base` R coding: read from inside the brackets {.build}

+ It can get messy very quickly. 

+ WORSE: You might not know what you are doing if you are not paying attention!

```{r}
plot(density(na.omit(eyData$edu_tertiary_pc_wkgpop), kernel = "gaussian"), col = "red")
```

## Piping ` %>% ` is about making your code readable to humans {.build}

+ **`base`**: `f(x)` means applying `f` on an object, `x`. 
+ **`magrittr`**: `x %>% f` means starting off with an object `x` and apply `f` upon it. 
+ Allows people to read codes from left to right.
+ Each function's option is within the righ bracket.

```{r}
mean(eyData$edu_tertiary_pc_wkgpop, na.rm = TRUE)

eyData$edu_tertiary_pc_wkgpop %>% 
  mean(na.rm = TRUE)
```


## The use of the `.` {.build}

```{r, echo = FALSE}
set.seed(1)
```

+ The `.` is a placeholder for everything before the pipe. 
+ It allows you to specify how and where the previous inputs should be used

```{r}
args(rnorm)
5 %>% rnorm(n = .)
5 %>% rnorm(n = 1, mean = .)
```


## The density plot revisited {.build}

+ It is **identical** to the previous density plot, but now it is much more readable for a human!

```{r}
eyData$edu_tertiary_pc_wkgpop %>% 
  na.omit %>% 
  density(kernel = "gaussian") %>% 
  plot(col = "red")
```

# `dplyr`


## Have you ever thought about `[,]`?

+ `base` R use `[,]` to subset data. 
+ But if you really think about it, why are we using the one symbol for two operations (**row**, **column**)?

```{r}
eyData[,c("id", "location")]
eyData[eyData$location == "AUS",]
```



## `dplyr` is a super way to manipulate your data {.build}

+ Think of subsetting rows and columns as two **separate different procedures**:
- `select` columns are operations on variables, and 
- `filter` rows are operations on observations

```{r}
eyData %>% select(id, location)
eyData %>% filter(location == "AUS")
```

## So what?? {.build}

+ Sure, you could stick with the `base` command. 

+ But sometimes, you realise that you could do something more powerful! (Notice how this would require a `for` loop or saving intermediate variables to achieve in `base`).

```{r}
eyData %>%
  filter(location == "AUS") %>%
  select(contains("population_ages"))
```

## `mutate` {.build}

+ How would you add in an extra column in a data?
+ And what if you want to add in a second column that is a transformation of the first column?

```{r, eval = FALSE}
eyData$gdp_aud_cap = eyData$gdp_usd_cap * 0.75
eyData$gdp_aud_cap_billions = eyData$gdp_aud_cap/1000
```

+ `mutate` allows you to do this in one function!

```{r}
eyData %>% 
  mutate(
    gdp_aud_cap = gdp_usd_cap*0.75,
    gdp_aud_cap_billions = gdp_aud_cap/1000
  )
```


## Summarise {.build}

+ Similar to `tapply`, but more flexible

```{r}
eyData %>% 
  group_by(location) %>% 
  summarise(
    medianQol = median(quality_of_life_measure),
    meanQol = mean(quality_of_life_measure)
  )

eyData %>% 
  group_by(location, time) %>% 
  summarise(medianQol = median(quality_of_life_measure))
```



# `ggplot2`

## I can't explain to you how to do a ggplot, you must learn it

+ *gg* stands for grammar of graphics. Which means, there is a consistent way to produce a graph. 

+ Advice through pain and tears: use `base` graphs if you can get it working very quickly (e.g. scatter plot or boxplot). Use `ggplot2` if your graph will inevitabily get complex. 

+ Legend!!! 

```{r}
eyData %>% 
  filter(location %in% c("AUS", "GBR", "SWE", "JPN")) %>%
  ggplot(aes(x = time,
             y = quality_of_life_measure,
             colour = gender)) +
  geom_point() + 
  geom_line() +
  facet_wrap(~location, nrow = 2)
```


# `tidyr`

## `tidyr` is a great companion with `ggplot2`

+ I know some of you used `reshape2::melt`. Which is a special case of `tidyr::gather`. The latter offers more control

```{r}
popAges = eyData %>% 
  select(location, contains("population_ages"))

popAges

popAges %>% 
  tidyr::gather(key = populationAgeGroups,
                value = percentage,
                -location)
```


## `separate` and `separate_rows` example: EY data dictionary
```{r}
dataDictionary = readxl::read_excel("EY-data-science-challenge-pack/Data Dictionary.xlsx")

dataDictionary[1,]

tmp = dataDictionary[1,] %>% 
  tidyr::separate_rows(col = DESCRIPTION, sep = "; ")

tmp

countryNameData = tmp %>% 
  tidyr::separate(col = DESCRIPTION, into = c("threeLetters", "countryName"), sep = " - ") %>% 
  dplyr::select(threeLetters, countryName)

countryNameData
```


## dplyr: Merging data

```{r}
eyData_joined = dplyr::left_join(eyData, 
                                 countryNameData, 
                                 by = c("location" = "threeLetters"))
eyData_joined
```



# `stringr`

## Not much I can say about this

+ The main criticism of base R string manipulation is the inconsistent arguments. 

```{r}
args(grep)
args(gsub)

args(stringr::str_detect)
args(stringr::str_replace_all)
```

